---
title: "log1p and expm1 operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{logexp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Rutils)
```

Two other operations that are common in machine learning research
and prone to numerical instability are computing
$\log(1+x)$ and $\exp(x)-1$ when $x$ is very small.
`Rutils` provides numerically-stable implementations of these two functions.

## Log1p

Suppose that $x=10^{-20}$.
This value is small but large enough to be represented in a computer.
However, since in floating-point arithmetic the numbers
we are able to represent in a computer are not evenly spaced,
the number $1+x=1+10^{-20}$ will be represented as $1$.
Hence naive computation of $\log(1+x)$
will return $\log(1)=0$.
The Log1p trick refers to the fact that, for small $x$,
$\log(1+x)\approx x$.
`Rutils::Log1p` checks if $x$ is too small for $1+x$ to be computed safely,
and if it is then it approximates $\log(1+x)$ with $x$.
See the example below:

```{r}
my_x <- 10^(-(10:20))
print(Log1p(my_x))
print(log(1+my_x))
```

For values of $x<10^{-16}$,
the naive computation returns $0$ but `Log1p` returns $x>0$.
Also note that `Log1p` is vectorized.


## Expm1

Again suppose that $x=10^{-20}$.
Then the operation $\exp(x)$ will return $0$
since the computer is not able to represent such a small number.
But using a Taylor's expansion,
$\exp(x)\approx 1+x$.
Again, we can approximate $\exp(x)-1$ with $x$ for sufficiently small $x$,
which `Rutils::Expm1` checks automatically.
See the example below:

```{r}
my_x <- 10^(-(10:20))
print(Expm1(my_x))
print(exp(my_x)-1)
```

As before, for $x<10^{-16}$ the naive computation returns 0
but `Expm1` returns $x>0$.
The function `Expm1` is also vectorized.
