---
title: "Rutils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to `Rutils`!
If you're tired of running into numerical issues when doing research,
or if you don't want to keep rewriting your `utils` script for each new project,
you've come to the right place!

Just install the package and then
```{r setup}
library(Rutils)
```
at the beginning of your script.

`Rutils` provides vectorized and numerically-stable implementations
of the LogSumExp, the Log1p, and the Expm1 tricks.

### Examples

Here are some quick examples,
but check the function vignettes for more!

```{r}
my_x <- -10^(10:20)

# naive computation fails:
log(sum(exp(my_x)))

# logsumexp trick does not:
logsumexp(my_x)
```


```{r}
my_x <- 1e-20

# naive computation fails:
log(1+my_x)

# Log1p trick does not:
Log1p(my_x)
```


```{r}
my_x <- 1e-20

# naive computation fails:
exp(my_x)-1

# Expm1 trick does not:
Expm1(my_x)
```
